---
title: "val_preprocessing_UC"
output: html_document
date: "2025-05-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
#install.packages("WGCNA")
#install.packages("gage")
#BiocManager::install("impute")
library(matrixStats)
library(WGCNA)
library(magrittr)
library(tibble)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(affy)
library(GEOquery)
library(limma)
library(umap)
library(affy)
library(vegan)
library(pheatmap)
library(openxlsx)
library(clusterProfiler)
library(org.Hs.eg.db)
#library(gage)
library(AnnotationDbi)
main_dir <- dirname(rstudioapi::getSourceEditorContext()$path) 
setwd(main_dir)
```

## **Информация о датасете:**

Patients with moderate-to-severe, active UC were treated with infliximab or placebo at weeks 0, 2, 6 and every 8 weeks thereafter. Biopsy samples (n=113) were collected from 48 patients at baseline, weeks 8 and 30 for RNA extraction and microarray analysis.

## Load the metadata for dataset

```{r message=FALSE, warning=FALSE}
gse97 <- getGEO(GEO = 'GSE23597', GSEMatrix = TRUE)
metadata97 <- pData(phenoData(gse97[[1]]))
```

### Choose useful columns

```{r}
# colnames(metadata97)
metadata_edited <- metadata97 %>%
  dplyr::select('title', 'geo_accession', 'source_name_ch1', 'characteristics_ch1.1',   'characteristics_ch1.2', 'characteristics_ch1.3', 'characteristics_ch1.4', 'characteristics_ch1.5') %>%
  dplyr::rename(tissue = 'source_name_ch1') %>%
  dplyr::rename(dose = 'characteristics_ch1.1') %>%
  mutate(dose = gsub("dose: ", "", dose)) %>%
  dplyr::rename(time_point = 'characteristics_ch1.2') %>%
  mutate(time_point = gsub("time: ", "", time_point)) %>%
  dplyr::rename(subject = 'characteristics_ch1.3') %>%
  mutate(subject = gsub("subject: ", "", subject)) %>%
  dplyr::rename(response_wk8 = 'characteristics_ch1.4') %>%
  mutate(response_wk8 = gsub("wk8 response: ", "", response_wk8)) %>%
  dplyr::rename(response_wk30 = 'characteristics_ch1.5') %>%
  mutate(response_wk30 = gsub("wk30 response: ", "", response_wk30)) %>%
  rownames_to_column(var = 'sample') %>%
  dplyr::rename(Geo_accession = 'geo_accession')

```



### Defining samples -- groups for comparison

The experiment involved patients who received placebo instead of infliximab. We select only those patients who received infliximab.

```{r}

# 1. EVERYTHING before treatment (Week zero)
all_beforeT <- metadata_edited %>%
  filter(dose != "placebo" & time_point == "W0")

# 2. EVERYTHING after 8 weeks of treatment
all_after_8W <- metadata_edited %>%
  filter(dose != "placebo" & time_point == "W8")
  
# 3. EVERYTHING after 30 weeks of treatment
all_after_30W <- metadata_edited %>%
  filter(dose != "placebo" & time_point == "W30")

# 4. Responders after 8 weeks before treatment
all_R8_beforeT <- metadata_edited %>%
  filter(dose != "placebo" & time_point == "W0" & response_wk8 == "Yes")
  
# 5. Responders after 30 weeks before treatment
all_R30_beforeT <- metadata_edited %>%
  filter(dose != "placebo" & time_point == "W0" & response_wk30 == "Yes")

# 6. Responders after 8 weeks after treatment
all_R8_afterT <- metadata_edited %>%
  filter(dose != "placebo" & time_point == "W8" & response_wk8 == "Yes")
  
# 7. Responders after 30 weeks after treatment
all_R30_afterT <- metadata_edited %>%
  filter(dose != "placebo" & time_point == "W30" & response_wk30 == "Yes")
  
# 8. All responders in general
all_afterT <- metadata_edited %>%
  filter(dose != "placebo" & (time_point == "W8" | time_point == "W30") & (response_wk30 == "Yes" | response_wk30 == "Yes"))

# 9. NONRESPONDERS BEFORE treatment
all_NR_beforeT <- metadata_edited %>%
  filter(dose != "placebo" & time_point == "W0" & response_wk30 == "No")

# 10. NONRESPONDERS AFTER treatment
all_NR_afterT <- metadata_edited %>%
  filter(dose != "placebo" & (time_point == "W8" | time_point == "W30") & response_wk30 == "No")
```

## Loading raw data
```{r message=FALSE, warning=FALSE, paged.print=FALSE}

# get supplementary files
getGEOSuppFiles("GSE23597", baseDir = '../../../data/UC/validation')

# untar files
untar("../../../data/UC/validation/GSE23597/GSE23597_RAW.tar", exdir = '../../../data/UC/validation/GSE23597/GSE23597_data')

raw.data <- ReadAffy(celfile.path = "../../../data/UC/validation/GSE23597/GSE23597_data")
```

### RMA - Normalization

```{r message=FALSE, warning=FALSE}
normalized.data <- rma(raw.data)
normalized.expr <- as.data.frame(exprs(normalized.data))
```

### Adding a Gene column.Symbol

And deleting lines without annotations

```{r message=FALSE, warning=FALSE}
gse_gs <- getGEO("GSE23597", GSEMatrix = TRUE)
feature.data <- gse_gs$GSE23597_series_matrix.txt.gz@featureData@data
feature.data <- feature.data[, c(1,11)]
feature.data <- feature.data %>%
  dplyr::rename(Gene.Symbol = "Gene Symbol")

normalized.expr <- normalized.expr %>%
  rownames_to_column(var = "ID")

gene_normalized.expr <- inner_join(normalized.expr, feature.data, by = "ID")


filtered_gene_normalized.expr <- gene_normalized.expr %>%
  filter(Gene.Symbol != "" & !is.na(Gene.Symbol))
```

### Grouping by Gene.Symbol and calculation of the average expression value

```{r message=FALSE, warning=FALSE}
mean_norm.expr <- filtered_gene_normalized.expr %>%
  group_by(`Gene.Symbol`) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
```

```{r}
# format the expression table
mean_norm.expr <- as.data.frame(mean_norm.expr)
rownames(mean_norm.expr) <- mean_norm.expr$Gene.Symbol
colnames(mean_norm.expr) <- gsub(".CEL.gz", "", colnames(mean_norm.expr))

mean_norm.expr <- subset(mean_norm.expr, select = -Gene.Symbol)
```

```{r}
# let's save patients before treatment
W0_ids <- all_beforeT$Geo_accession

W0_ids <- W0_ids[W0_ids %in% colnames(mean_norm.expr)]

W0_expr_data <- mean_norm.expr[ , W0_ids]
```


```{r}
# Make sure that the order of the patients in the expression and metadata match
W0_ids <- colnames(W0_expr_data)
all_beforeT_matched <- all_beforeT[match(W0_ids, all_beforeT$Geo_accession), ]

# Let's check the match:
stopifnot(all(W0_ids == all_beforeT_matched$Geo_accession))

# Creating a vector: 1 = responder, 0 = non-responder
response_vector <- ifelse(all_beforeT_matched$response_wk8 == "Yes", 1, 0)

```

```{r}
hub_genes <- c("IL6", "TLR4", "TLR8", "IL1B", "CXCL8",
               "TLR2", "ITGAM", "PTGS2", "S100A12", "FCGR3B")

ref_gene <- "ACTB"

```

```{r}
genes_all <- c(hub_genes, ref_gene)
expr_subset <- W0_expr_data[rownames(W0_expr_data) %in% genes_all, ]

```


```{r}
response_vector_val <- response_vector
expr_hub_genes_val <- expr_subset
write.csv(expr_hub_genes_val, file = "../../../tables/UC/validation/expr_hub_genes_val.csv", row.names = TRUE)

write.csv(data.frame(response = response_vector_val), file = "../../../tables/UC/validation/response_vector_val.csv", row.names = FALSE)
```